Neural models for target-based computer-assisted musical orchestration: a preliminary study

### 2020-05-16 paper/latex/paper.tex
fixed typos - final version submitted for review

%
\maketitle
%
\begin{abstract}
In this paper we will do a preliminary exploration on how neural networks can be used for the task of target-based computer-assisted musical orchestration. We will show how it possible to model this  musical problem as a classification task and we will propose two deep learning models. We will show, first, how they perform as classifiers for musical instruments recognition by comparing them with specific baselines. We will then show how they perform, both qualitative and quantitative, in the task of computer-assisted orchestration by comparing them with state-of-the-art systems. Finally, we will highlight benefits and problems of neural approaches for assisted orchestration and we will propose possible future steps.
\end{abstract}
%
\section{Introduction}\label{sec:introduction}

The development of computational tools to assist and inspire the musical composition process constitutes an import research area known as \emph{Computer-Assisted Composition (CAC)} \cite{FerVic2013, Ari2005}. Within CAC, target-based computer-assisted orchestration is an important example how machine learning can be used for {enhancing} and {assisting} music creativity \cite{Maresz2003}.

@@

\subsection{Form orchestration to classification}

@@

Three baseline classifiers have been trained to recognize either the instrument only for a total of 12 classes or the pair instrument-pitch \emph{class} (a pitch class is the pitch of a note without the corresponding octave, e.g. C, D\#, G, etc.) for a total of 25 classes.

Two neural classifiers, instead, have been trained on the more complex problem of recognizing instrument-pitch pairs, for a total of 424 classes (i.e. the neural classifiers would classify instruments and pitches with respective octaves). We want to point out, however, that this is still a simplified problem compared to assisted orchestration. Even in the case of instrument-pitch pair classification, indeed, we would miss the dynamics of the note. To cope with this problem, in the orchestration phase, we quantize the probability functions generated by the models. When a network is given a target sound as input, it will attempt to apply the same classification rules, outputting a vector of the probabilities of each sample being in the target sound. Since the classes only encode instrument-pitch pairs and since we also need the dynamics, we quantize the probability function generated by the network into three echelons coresponding, respectively, to the dynamic levels \emph{pianissimo}, \emph{mezzoforte} and \emph{fortissimo} (see section \ref{sec:orchestration}). By taking the instruments and pitches that have the highest probability and by using quantized probabilities to retrieve dynamics, a full orchestrated solution can be created.  Figure \ref{fig:method} illustrates the described ideas.

@@

The choice of using the Mel spectrogram and MFCCs as input features for classification models is common in music information retrieval \cite{McKinney2003} and can be considered to be an accurate representation of sound and musical signals. The Mel spectrogram is generated using an FFT hop length of 2048 samples (e.g. the window of each FFT was $46$ms wide), and a number of Mel bins equal to 128. Therefore, the Mel features fed to the model were matrices of size $128\times 345$. We used the Librosa package in Python to compute the features; more details on the exact computations can be found in \cite{mcfee15}. The choice of the hop length was made by doing a compromise between the amount of information extracted by each FFT window, and the ability to capture changes in the dynamic, assuming that no change faster than $10$ms would be allowed.

@@

\subsection{Baseline}
The baseline classifiers that we tested are support vector machine (SVM), random forest (RF), and K-nearest neighbours (KNN). We used the implementations provided in the \emph{scikit-learn }library for Python \cite{scikit-learn}. More specifically, for SVM we used a non-linear RBF kernel and for RF we set the maximum depth of each tree to be 15.  All of the baseline experiments used 50,000 generated samples with a train-test split of 60/40. Each sample was made by a combination of one to ten instruments and was four seconds in length. 
In this case, differently from the neural models, the features used are the MFCCs of the resulting combination; we chose to use MFCCs for this setting, instead of the Mel spectrogram, to have a number of features that is more appropriate for the parametric classifiers. We found SVM to have the highest accuracy of the three classifiers across all experiments.

@@

To better approximate our original problem of identifying instrument and pitch, however, we performed experiments in which two instruments were selected and for input data that contained samples from one of those two instruments, both the instrument and pitch \emph{class} of the sample would be classified. The pitch class is the note name without the octave, e.g. C, D\#, G, etc. The input was a combination of two instruments drawn from a possible twelve instruments, and the classifier attempted to identify which instruments were present and for the specified instruments, say Violin and Viola, which pitch classes were present. If another instrument was present in the input combination that was not Violin or Viola, the classifier would simply identify that an instrument that was not one of the two was present. The accuracies from this experiment are outlined in table \ref{tab:baselines}. Since RF performed worse than SVM is every experiment, we stopped testing with it and used SVM from that point on.

We then performed this same experiment with three instruments having their pitch class identified. Flute, Oboe, and Violin reached an accuracy of 11.1\%, and Bass Tuba, Trumpet, and Trombone was 0.5\%. As we increased the number of instruments whose pitch classes was being identified, the accuracy continued to drop. For classifying the pitch class of four instruments: Oboe, French Horn, Violin, and Flute, the accuracy was 2.7\%.

@@

The first deep model we trained as a classifier for musical instruments and pitches was a CNN with a LSTM unit. CNNs show good performance on classification problems for their ability to extract spatial features, and have shown success in audio classification \cite{Hershey16}. LSTM units provide a way to learn long term dependencies in the data \cite{Hochreiter97}, which is relevant given the sequential nature of audio.

@@

Different experiments were made by varying the number $N$ of samples in each mixture sample. We used an orchestra of 10 instruments containing Horn, Oboe, Violin, Viola, Cello, Flute, Trombone, Bassoon, Trumpet in C and Clarinet in Bb.

@@

After training the neural models for classification, we finally tested them for the task of target-based computer-assisted orchestration. For this purpose, we decided to use the models trained on combinations of 10 instruments: French Horn, Oboe, Violin, Viola, Cello, Flute, Trombone, Bassoon, Trumpet, and Clarinet. The total number of classes in this setting is 424, each one representing an instrument-pitch pair.

@@

Since we decided not to train our models to classify the dynamics of a sample (despite TinySOL having \emph{pianissimo}, \emph{mezzoforte}, and \emph{fortissimo} recordings), the dynamics is determined by the probability of each sample as output by the model. If the model outputs a probability higher than $0.66$ for a sample, the fortissimo version of the sample is used. If the probability is between $0.33$ and $0.66$, then the mezzoforte version is selected and if it is less than $0.33$ the pianissimo version is used. The idea behind this quantization, is that samples that are the most represented in the target should appear as the loudest in the orchestrated solution.

@@

Orchidea implements many advanced features that are not supported by our models. For example, it is able to apply symbolic constraints to the search, hence allowing only specific instrumental combinations or playing styles. It is also able to reduce the search space by applying harmonic analysis on the target sound.
The dominant harmonic partials of the target are identified, and the search space is limited to only include samples of those pitches. For example, if the target is a recording of an instrument playing a C4, then the partials identified may be C4, C5, G5, and E6. The model would then only consider samples of these pitches to be used in the solution. This leads to a solution whose harmonics are much closer to the target, which is an important part of aural similarity.
Orchidea's solutions, moreover, can use any number of instruments included in the orchestra specified by the user, thus having variable-sized orchestrations from a single instrument to the whole set (this propriety is called \emph{sparsity} of the solution).

@@

We calculated the distance between the target samples and our orchestrated solutions. We then orchestrated the same targets with Orchidea and calculated the distance for Orchidea's solutions. A comparison of these results is in Tab. \ref{orch_eval}.

@@

The first natural extension would be to support sparsity in our models. Our current models, indeed, orchestrate all targets using a constant number of instruments and it is not able to drop specific instruments from the solution. This does not take into account the density of different targets. By using sparse solutions, in which the model decides how many samples should be used to best represent the target, would allow a small number of samples to be used for sonically sparse sounds and many to be used for sonically dense sounds. This would generate more meaningful orchestrations that would compare more favourably to the state of the art.

@@

\bibitem{FerVic2013} J.D. Fern\'andez and F. Vico. {AI Methods in Algorithmic Composition: A Comprehensive Survey}. Journal of Artificial Intelligence Research, vol.48, pp. 513--582, 2013.

\bibitem{Ari2005} C. Ariza, {Navigating the landscape of computer aided algorithmic composition systems: A definition, seven descriptors, and a lexicon of systems and research}. In Proceedings of the International Computer Music Conference, 2005.

@@

\bibitem{Cella2020}
Carmine~Emanuele Cella.
\newblock Orchidea: a comprehensive framework for target-based assisted orchestration.
\newblock In preparation; available on request from the author, 2020.

\bibitem{Cella2020b}
Carmine~Emanuele Cella, Daniele Ghisi, Vincent Lostanlen, Fabi\'en Levy, Yan Maresz and Joshua Fineberg.
\newblock OrchideaSOL: a dataset of extended instrumental techniques for computer-aided orchestration.
\newblock ICMC (under review), 2020.

@@

\bibitem{scikit-learn}
F.~Pedregosa, G.~Varoquaux, A.~Gramfort, V.~Michel, B.~Thirion, O.~Grisel,
  M.~Blondel, P.~Prettenhofer, R.~Weiss, V.~Dubourg, J.~Vanderplas, A.~Passos,
  D.~Cournapeau, M.~Brucher, M.~Perrot, and E.~Duchesnay.
\newblock Scikit-learn: Machine learning in {P}ython.
\newblock {\em Journal of Machine Learning Research}, 12:2825--2830, 2011.

>>>>

%
\maketitle
%
\begin{abstract}
In this paper we will do a preliminary exploration on how neural networks can be used for the task of target-based computer-assisted musical orchestration. We will show how it is possible to model this  musical problem as a classification task and we will propose two deep learning models. We will show, first, how they perform as classifiers for musical instrument recognition by comparing them with specific baselines. We will then show how they perform, both qualitatively and quantitatively, in the task of computer-assisted orchestration by comparing them with state-of-the-art systems. Finally, we will highlight benefits and problems of neural approaches for assisted orchestration and we will propose possible future steps.
\end{abstract}
%
\section{Introduction}\label{sec:introduction}

The development of computational tools to assist and inspire the musical composition process constitutes an import research area known as \emph{Computer-Assisted Composition (CAC)} \cite{FerVic2013, Ari2005}. Within CAC, target-based computer-assisted orchestration is an important example of how machine learning can be used for {enhancing} and {assisting} music creativity \cite{Maresz2003}.

@@

\subsection{From orchestration to classification}

@@

Three baseline classifiers have been trained to recognize either the instrument only for a total of 12 classes (the instruments in TinySOL) or the pair instrument-pitch \emph{class} (a pitch class is the pitch of a note without the corresponding octave, e.g. C, D\#, G, etc.) for a total of 25 classes (we will classifiy combinations of 2 instruments for 12 pitch classes each, plus one extra \emph{trash} class, see section \ref{sec:baseline}).

Two neural classifiers, instead, have been trained on the more complex problem of recognizing instrument-pitch pairs, for a total of 424 classes (i.e. the neural classifiers would classify instruments and pitches with respective octaves, hence 12 instruments for 12 pitch classes and three dynamics; note that not all the samples in TinySOL have all dynamics). We want to point out, however, that this is still a simplified problem compared to assisted orchestration. Even in the case of instrument-pitch pair classification, indeed, we would miss the dynamics of the note. To cope with this problem, in the orchestration phase, we quantize the probability functions generated by the models. When a network is given a target sound as input, it will attempt to apply the same classification rules, outputting a vector of the probabilities of each sample being in the target sound. Since the classes only encode instrument-pitch pairs and since we also need the dynamics, we quantize the probability function generated by the network into three echelons coresponding, respectively, to the dynamic levels \emph{pianissimo}, \emph{mezzoforte} and \emph{fortissimo} (see section \ref{sec:orchestration}). By taking the instruments and pitches that have the highest probability and by using quantized probabilities to retrieve dynamics, a full orchestrated solution can be created.  Figure \ref{fig:method} illustrates the described ideas.

@@

The choice of using the Mel spectrogram and MFCCs as input features for classification models is common in music information retrieval \cite{McKinney2003} and can be considered to be an appropriate representation of sound and musical signals. The Mel spectrogram is generated using an FFT hop length of 2048 samples (the window of each FFT was $46$ms wide), and a number of Mel bins equal to 128. Therefore, the Mel features fed to the model were matrices of size $128\times 345$. We used the Librosa package in Python to compute the features; more details on the exact computations can be found in \cite{mcfee15}. The choice of the hop length was made by doing a compromise between the amount of information extracted by each FFT window, and the ability to capture changes in the dynamic, assuming that no change faster than $10$ms would be allowed.

@@

\subsection{Baselines}
\label{sec:baseline}

In order to get a betters sense on the complexity of the problem, we tested three baseline classifiers: support vector machine (SVM), random forest (RF), and K-nearest neighbours (KNN). We used the implementations provided in the \emph{scikit-learn }library for Python \cite{scikit-learn}. More specifically, for SVM we used a non-linear RBF kernel and for RF we set the maximum depth of each tree to be 15.  All of the baseline experiments used 50,000 generated samples with a train-test split of 60/40. Each sample was made by a combination of one to ten instruments and was four seconds in length. 
In this case, differently from the neural models, the features used are the MFCCs of the resulting combination; we chose to use MFCCs for this setting, instead of the Mel spectrogram, to have a lower number of features, that is more manageable by parametric classifiers. We found SVM to have the highest accuracy of the three classifiers across all experiments.

@@

To better approximate our original problem of identifying instrument and pitch, however, we performed experiments in which two instruments were selected and for input data that contained samples from one of those two instruments, both the instrument and pitch \emph{class} of the sample would be classified. The pitch class is the note name without the octave, e.g. C, D\#, G, etc. The input was a combination of two instruments drawn from a possible twelve instruments, and the classifier attempted to identify which instruments were present and for the specified instruments, say Violin and Viola, which pitch classes were present. If another instrument was present in the input combination that was not Violin or Viola, the classifier would simply identify that an instrument that was not one of the two was present. The accuracies from this experiment are outlined in Table \ref{tab:baselines}. Since RF performed worse than SVM in every experiment, we stopped testing with it and used SVM from that point on.

We then performed this same experiment with three instruments having their pitch class identified. Flute, Oboe, and Violin reached an accuracy of 11.1\%, and Bass Tuba, Trumpet, and Trombone was 0.5\%. As we increased the number of instruments whose pitch classes was being identified, the accuracy continued to drop.Indeed, for classifying the pitch class of four instruments (Oboe, French Horn, Violin, and Flute) the accuracy was 2.7\%.

@@

The first deep model we trained as a classifier for musical instruments and pitches was a CNN with a LSTM unit.

%CNNs show good performance on classification problems for their ability to extract spatial features, and have shown success in audio classification \cite{Hershey17}. LSTM units provide a way to learn long term dependencies in the data \cite{Hochreiter97}, which is relevant given the sequential nature of audio.

@@

Different experiments were made by varying the number $N$ of samples in each mixture. We used an orchestra of 10 instruments containing Horn, Oboe, Violin, Viola, Cello, Flute, Trombone, Bassoon, Trumpet in C and Clarinet in Bb.

@@

After training the neural models for classification, we finally tested them for the task of target-based computer-assisted orchestration. For this purpose, we decided to use the models trained on combinations of 10 instruments, namely French Horn, Oboe, Violin, Viola, Cello, Flute, Trombone, Bassoon, Trumpet, and Clarinet. Each class, in this model, represents an instrument-pitch pair.

@@

Since we decided not to train our models to classify the dynamics of a sample (despite TinySOL having \emph{pianissimo}, \emph{mezzoforte}, and \emph{fortissimo} recordings), the dynamics are determined by the probability of each sample as output by the model. If the model outputs a probability higher than $0.66$ for a sample, the fortissimo version of the sample is used. If the probability is between $0.33$ and $0.66$, then the mezzoforte version is selected and if it is less than $0.33$ the pianissimo version is used. The idea behind this quantization is that samples that are the most represented in the target should appear as the loudest in the orchestrated solution.

@@

Orchidea implements many advanced features that are not supported by our models. For example, it is able to apply symbolic constraints to the search, hence allowing only specific instrumental combinations or playing styles. It is also able to reduce the search space by applying harmonic analysis on the target sound.
The dominant harmonic partials of the target are identified, and the search space is limited to only include samples of those pitches. For example, if the target is a recording of an instrument playing a C4, then the partials identified may be C4, C5, G5, and E6. The model would then only consider samples of these pitches to be used in the solution. This leads to a solution whose harmonics are much closer to the target, which is an important part of aural similarity.
Orchidea's solutions, moreover, can use any number of instruments included in the orchestra specified by the user, thus having variable-sized orchestrations from a single instrument to the whole set (this property is called \emph{sparsity} of the solution).

@@

We calculated the distance between the target samples and our orchestrated solutions. We then orchestrated the same targets with Orchidea and calculated the distance for Orchidea's solutions. A comparison of these results is in Table \ref{orch_eval}.

@@

The first natural extension would be to support sparsity in our models. Our current models, indeed, orchestrate all targets using a constant number of instruments and it is not able to drop specific instruments from the solution. This does not take into account the density of different targets. Sparse solutions, in which the model decides how many samples should be used to best represent the target, would allow a small number of samples to be used for sonically sparse sounds and many to be used for sonically dense sounds. This would generate more meaningful orchestrations that would compare more favourably to the state of the art.

@@

\bibitem{FerVic2013} Jose David Fern\'andez and Francisco Vico. {AI Methods in Algorithmic Composition: A Comprehensive Survey}. Journal of Artificial Intelligence Research, vol.48, pp. 513--582, 2013.

\bibitem{Ari2005} Christopher Ariza, {Navigating the landscape of computer aided algorithmic composition systems: A definition, seven descriptors, and a lexicon of systems and research}. In Proceedings of the International Computer Music Conference, 2005.

@@

\bibitem{Cella2020}
Carmine-Emanuele Cella.
\newblock Orchidea: a comprehensive framework for target-based assisted orchestration.
\newblock In preparation; available on request from the author, 2020.

\bibitem{Cella2020b}
Carmine-Emanuele Cella, Daniele Ghisi, Vincent Lostanlen, Fabi\'en Levy, Yan Maresz and Joshua Fineberg.
\newblock OrchideaSOL: a dataset of extended instrumental techniques for computer-aided orchestration.
\newblock ICMC (under review), 2020.

@@

\bibitem{scikit-learn}
Fabian Pedregosa, Ga\"el Varoquaux, Alexandre
Gramfort, Vincent Michel, Bertrand Thirion, Olivier
Grisel, Mathieu Blondel, Peter Prettenhofer, Ron
Weiss, Vincent Dubourg, Jake Vanderplas, Alexandre
Passos, David Cournapeau, Matthieu Brucher,
Matthieu Perrot, \'Edouard Duchesnay.
\newblock Scikit-learn: Machine learning in {P}ython.
\newblock {\em Journal of Machine Learning Research}, 12:2825--2830, 2011.
